---
title: "Running models and diagnostics"
author: "Andrew Parnell"
date: "28 June 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Welcome to Practical 2: running models and diagnostics. In this practical we will:

- Work through some ready-to-run R scripts
- Learn how to run regression and classification models
- Learn how to run diagnostics and check fit

## Instructions

This document is much shorter than the previous practical. To run this practical properly you need to install the following two packages:
```{r, eval=FALSE}
install.packages('pROC')
install.packages('ROCR')
```

The things to do are:

1. Copy or download the two code files for the [prostate](https://raw.githubusercontent.com/andrewcparnell/bio_course/master/code/prostate_code.R) example and the [heart disease](https://raw.githubusercontent.com/andrewcparnell/bio_course/master/code/SA_code.R) example into new R scripts
2. Start with the prostate code and work through the script, then move on to the heart disease code.
3. Try to understand what each line of code is doing (use the help commands to look up functions you don't understand)
4. At various places in the code are questions or other tasks you can complete if you want to
5. If you get stuck please raise your hand and I will try to help
